---
title: "Principles of Data Science"
author: "Oghenefejiro Jennifer Imene"
date: "2023-11-30"
output:
  word_document: default
  pdf_document: default
student ID: "00693014"
---


```{r "loading the package"}
#Load the appropriate libraries 
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(psych)
library(scales)
library(ggthemes)
library(flextable)
library(xtable)
library(gt)
library(moments)
library(xtable)
library(corrplot)
library(car)
library(tseries)
library(TTR)
library(forecast)
```



## Reseach Objective:


The objective of this research is to cross examine the changes in population 
and life expectancy for top 12 richest countries in Europe

Population is the total number of people or inhabitants in a country or region.

```{r}

#load in the data

birth_data <- read.csv("Births.csv")
Death_data <- read.csv("Deaths.csv")
life_data  <- read.csv("Life expectancy.csv")
Migration_data <- read.csv("Migration.csv")
population_data <- read.csv("Total population.csv")


```


```{r}
#rename the cols
birth_data <- birth_data %>%  select  (-Variant) %>%  rename ( `number of birth` = Value)  

Death_data <- Death_data  %>% select  (-Variant) %>%  rename ( `number of deaths` = Value)   

life_data  <- life_data  %>% select  (-Variant) %>% rename (  `life expectancy` = Value)  

Migration_data <- Migration_data %>% select  (-Variant) %>% rename(`number of Migration` = Value)  

population_data <- population_data %>% select  (-Variant) %>%  rename (`Total population` = Value) 

```




```{r}
#combining all the data sets into one variable

data <- cbind(birth_data,Death_data, life_data,Migration_data,population_data)


```


```{r}

#drop unwanted columns
data <- data %>%  select(-4,-5,-7,-8,-10,-11,-13,-14)

head(data)
colnames(data)

```


```{r}
#Describe the data
describe(data)

```

```{r}
#convert the country var into a factor

data$Country <- as.factor(data$Country)

```


```{r}

data_stats <-  data %>% select(-2) %>% pivot_longer(!Country, names_to = "Indicators", values_to  = "Rates")
#make some conversions

Indicator <- factor(data_stats$Indicators,
               labels = c("Birthnumber", "Death number", "life expectancy number", "Migration number","total Population"))

Rate <- data_stats$Rates


`Average value` <- tapply( Rate,Indicator, mean)
`Median Value` <- tapply( Rate,Indicator,median)
 Mode <- tapply( Rate,Indicator,max)
`Standard Deviation` <- tapply( Rate,Indicator,sd)
`Skewness` <- tapply( Rate,Indicator,skewness)
Kurtosis <- tapply( Rate,Indicator,kurtosis)

#predictor_stats = rbind(Indicator_averages,Indicator_median,Indicator_mode ,Indicator_standard_deviation,Indicator_skewness,Indicator_kurtosis)
predictor_stats = rbind(`Average value`,`Median Value`, Mode ,`Standard Deviation`,`Skewness`,Kurtosis)

View(predictor_stats)
#predictor_stats

```


```{r}
#Descriptives visualizatons
#Box plots (Inicators by Country)

ggplot(data, aes(x = Country, y = `number of birth`, fill = Country))+ geom_boxplot()  +ggtitle(labs("Birth Rates boxplots by country"))

ggplot(data, aes(x = Country, y = `number of deaths`, fill = Country))+ geom_boxplot()  +ggtitle(labs("Death Rates boxplots by country"))

ggplot(data, aes(x = Country, y = `life expectancy`, fill = Country))+ geom_boxplot() + ggtitle(labs("Life expectancy values boxplots by country"))

ggplot(data, aes(x = Country, y = `number of Migration`, fill = Country))+ geom_boxplot()  +ggtitle(labs("Migration Rates boxplots by country"))

ggplot(data, aes(x = Country, y = `Total population`, fill = Country))+ geom_boxplot()  +ggtitle(labs("Population Rates boxplots by country"))



```

```{r}
#trend of the indicators per country
#birth rates

ggplot(data=data, aes(x=Year, y=`number of birth`,color=Country,group=Country))  + geom_line(size=2, alpha=0.5, linetype=1) +
 ylab("number of birth")  + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Trend of birth numbers in top 12 richest countries in Europe")

#death rates
ggplot(data=data, aes(x=Year, y=`number of deaths`,color=Country,group=Country))  + geom_line(size=2, alpha=0.5, linetype=1) +
 ylab("number of deaths")  + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Trend of death numbers in top 12 richest countries in Europe")

#fertility rates
ggplot(data=data, aes(x=Year, y=`life expectancy`,color=Country,group=Country))  + geom_line(size=2, alpha=0.5, linetype=1) +
 ylab("life expectancy")  + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Trend of life expectancy in top 12 richest countries in Europe")

#migration rates
ggplot(data=data, aes(x=Year, y=`number of Migration`,color=Country,group=Country))  + geom_line(size=2, alpha=0.5, linetype=1) +
 ylab("number of Migration")  + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Trend of migration numbers in top 12 richest countries in Europe")

#population rates
ggplot(data=data, aes(x=Year, y=`Total population`,color=Country,group=Country))  + geom_line(size=2, alpha=0.5, linetype=1) +
 ylab("Total population")  + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Trend of Total population in top 12 richest countries in Europe")

```


```{r}
#Descriptive stats
#Compute averages of the indicators by country

#Average birth number

Average_Birth_number <-  data %>% 
               group_by(Country) %>%
                summarise( average_birth_number = mean(`number of birth`))


 


#Average death number

Average_Death_number <-  data %>% 
               group_by(Country) %>%
                summarise( average_death_number = mean(`number of deaths`))

 
#Average migration number

Average_Mig_number <-  data %>% 
               group_by(Country) %>%
                summarise( average_migration_number = mean(`number of Migration`))

  
 
#Average fertility rate

Average_life_expectancy <-  data %>% 
               group_by(Country) %>%
                summarise( average_life_expectancy = mean(`life expectancy`))

 


#Average change in population rate
  
Average_Pop_number <-  data %>% 
               group_by(Country) %>%
                summarise( average_population_number = mean(`Total population`))


# ggplot(data = avg_birth_rate, aes(x = reorder(Country,average_birth_rate), y  = average_birth_rate, fill = Country)) + geom_bar(stat = "identity")
```


```{r}
#Consolidate the average rates of indicators into a table
#drop the repeated columns

figures <- cbind(Average_Birth_number,Average_Death_number,Average_life_expectancy,Average_Mig_number,Average_Pop_number) %>% 
#drop the repeated columns
  
  select (-3,-5,-7,-9)

#The Rates table
table1 <- figures %>%
  gt() %>%
  tab_header(title = "Average numbers of indicators per country") 

table1

```


```{r}
#Visualize the average rates of all indicators across countries
#first restructure the data

figures_reshape <- figures %>%
       pivot_longer(!Country, names_to = "Indicators", values_to  = "Values")

head(figures_reshape)
tail(figures_reshape)
```




##Hypothesis Testing

We shall be using hypothesis testing of means to determine if there is a significant difference in the average number of population  in two of countries from the data set. 

The countries of interests are France are Italy, as they have almost the same average population over the last 12 years.

First, we have to extract the data, then check to confirm the data follows a normal distribution

```{r}

data_xtraction <- function(df,countryname){
    df %>% filter( Country == countryname)
}

France <- data_xtraction(data,"France")

Italy <- data_xtraction(data,"Italy")

```


```{r}
#normality test

shapiro.test(France$`Total population`)
shapiro.test(Italy$`Total population`)

```
Interpretation:

Since the p-values for both test (0.2463, 0.1594) are both greater than 0.05, we have statistical reasons to conclude the data set follows a normal distribution. 

```{r}
#Hypothesis one  (forFrance)
France_average_pop <-mean(France$`Total population`)
Italy_average_pop <- mean(Italy$`Total population`)

#Set the hypothesis
```

Ho: France average population rate = 63805.68
H1: France average population != 63805.68

Ho: Italy average population rate = 59877.94
H1: Italy average population != 59877.94

```{r}
#The Tests

France_outcome <- t.test(France$`Total population`, mu = 63805.68)
Italy_outcome <- t.test(Italy$`Total population`, mu = 59877.94)
France_outcome
Italy_outcome

```

##Correlation Test for the predictors

Correlation measures the linear association between two variables.
Before correlation test is performed, the data set must meet some certain criteria. They are:

1. Level of Measurement: The two variables should be measured at the interval or ratio level.

2. Linear Relationship: There should exist a linear relationship between the two variables.

3. Normality: Both variables should be roughly normally distributed.

4. Related Pairs: Each observation in the dataset should have a pair of values.

5. No Outlines: There should be no extreme outliers in the dataset.




We shall be lookng at the relationship between the 5 indicators

```{r}

#correlation
#get the correlation of the indicators

cor_data <- data %>% select (3,4,5,6,7)

corrplot(cor(cor_data), method = "number", type = "upper")

```

##Fit a regression Model
 We would be fixing two regression model for 2 out of the 12 countries in terms of their total population. The 2 countries are: Norway and Switzerland.
 The reason these countries were chosen, is because they are the countries with the least population numbers. 
 
## The objective of the regression model:
To get a model that can forecast the population number for the next few years 

```{r}

#extract the data 

norway <- data_xtraction(data,"Norway")
  
switzerland <- data_xtraction(data,"Switzerland")

```



```{r}

#Get the correlation plots

norway <- norway %>% select (-1,-2)
switzerland <- switzerland %>% select (-1,-2)

corrplot(cor(norway), method = "number", type = "upper", main = "Correlation plot for norway")
corrplot(cor(switzerland), method = "number", type = "upper", main= "Correlation plot forswitzerland")

```


```{r}

#Get the model
norway_model <- lm(formula = `Total population` ~ `number of birth` + `number of deaths` + `number of Migration` , data = norway)

switzerland_model <- lm(formula = `Total population` ~ `number of birth` + `number of deaths` + `number of Migration` , data = switzerland)

summary(norway_model)
summary(switzerland_model)

```


###Interpretation

A look at two models revealed that all coefficients are statistically significant at 0.05 level, except for number of deaths and number of Migration
in the case of norway which is not significant. 

The equation for Norway will be :

Population Rate = 8752.784  -50.310 x number of births  


The equation for switzerland will be :

Population Rate = 2323.145 + 57.083  x number of births  22.039 x number of deaths -4.943 x number of Migration


Adjusted R^^2 is a corrected goodness-of-fit (model accuracy) measure for linear models. It identifies the percentage of variance in the response variable (Population rate) that is explained by the Independent variables. 

R^^2 are 1 in both models. This value indicates that the models are a good fit and they have higher prediction power. 



###Check if model meets assumptions

1. Linearity Check

```{r}
##Ensure the model fits the model
# linearity test
# scatter plot

norway_df <- norway

attach(norway_df)
pairs(~ `Total population` + `number of birth` + `number of deaths` + `number of Migration` ,  main= "scatter plot for Norway",lower.panel = NULL,  col= c('red','green'))


switzerland_df <- switzerland


attach(switzerland_df)
pairs(~`Total population` + `number of birth` + `number of deaths` + `number of Migration` , main= "scatter plot for Switzerland",lower.panel = NULL,  col= c('red','green'))

```


2. Residuals’ Independence
```{r}

plot(norway_model, 1)
plot(switzerland_model, 1)


```


3. Normality of residuals

```{r}

plot(norway_model, 2)
plot(switzerland_model, 2)

```

4. Equal variances of the residuals (Homoscedasticity)

```{r}
plot(norway_model, 3)
plot(switzerland_model, 3)
```


5. No multicollinearity
```{r}
vif(norway_model)
vif(switzerland_model)
```
Remove the variables highly correlated in the norway_model  



## Time series Analysis
For the time series analysis, we shall be using the same countries we used for the regression models- Norway & Switzerland


```{r}
#build the ts data

norway_ts_d <- norway$`Total population`
swit_ts_d  <- switzerland$`Total population`

norway_ts <- ts(norway_ts_d, frequency = 1 , start  = c(2011,1))
swit_ts <- ts(swit_ts_d,frequency  =1, start = c(2011,1))
```


```{r}
#timeseries plot
plot.ts(norway_ts,col = "red")
plot.ts(swit_ts,col = "blue")

```


```{r}
#test for data stationarity

adf.test(norway_ts)
adf.test(swit_ts)

```
The p-value for both test are greater than 0.05, we fail to reject the null hypothesis.

This means the time series is non-stationary. 

Next is to make the time series data by differencing it.

```{r}
#difference the series

df_norway_ts <- diff(norway_ts, difference = 2)
df_swit_ts <- diff(swit_ts, differences = 1)
plot(df_norway_ts)
plot(df_swit_ts)
  
```


If your time series is stationary, or if you have transformed it to a stationary time series by differencing d times, the next step is to select the appropriate ARIMA model, which means finding the values of most appropriate values of p and q for an ARIMA(p,d,q) model. To do this, you usually need to examine the correlogram and partial correlogram of the stationary time series.

```{r}
acf(df_norway_ts, lag.max=20)
pacf(df_norway_ts, lag.max=20)
acf(df_swit_ts, lag.max=20)
pacf(df_swit_ts, lag.max=20)
```




```{r}
#plot the models ( For norway)
norway_arima <- arima(norway_ts, order=c(1,2,0))
norway_arima
norway_arima_forecast <- forecast(norway_arima, h=5)


switz_arima <- arima(swit_ts, order=c(1,0,1))
switz_arima
switz_arima_forecast <- forecast(switz_arima, h=5)
```


```{r}
#The forecasts
 norway_arima_forecast
switz_arima_forecast

plot(norway_arima_forecast)
plot(switz_arima_forecast)
```


```{r}
#residuals check

acf(norway_arima_forecast$residuals)
Box.test(norway_arima_forecast$residuals, type="Ljung-Box")

acf(switz_arima_forecast$residuals)
Box.test(switz_arima_forecast$residuals, type="Ljung-Box")

```






```

